---
title: "01 - Downloading Data - Census"
output: html_document
---


This script uses the Census API to download a number of demographic variables over multiple years at the Census-tract level from the American Community Survey (five-year estimates).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source('functions/f03b--data--census.R')
```


The demographic variables selected are those used in one of my projects, and they are also commonly used in the literature on neighborhoods and crime (see, for example, Peterson & Krivo 2010, Divergent Social Worlds).


```{r kbl_vars, echo=FALSE}
tribble(~Variable,      ~Description,
        'pop.t',        'Total residential population',
        'pop_1_plus.t', 'Total residents, ages ≥ 1',
        'pop_25_64.t',  'Total residents, ages 25-64',
        'famhh.t',      'Total family households w/ children < 18',
        'hh.t',         'Total households',
        'pop_black',    'Black residential population',
        'pop_latinx',   'Latinx residential population',
        'males_15_17y', 'Male residents, ages 15-17',
        'males_18_19y', 'Male residents, ages 18-19',
        'males_20y',    'Male residents, age 20',
        'males_21y',    'Male residents, age 21',
        'males_22_24y', 'Male residents, ages 22-24',
        'males_25_29y', 'Male residents, ages 25-29',
        'same_home_1y', 'Residents living in same home 1 yr ago',
        'no_hsd_25_64', 'Residents (25-64) w/o high school diploma',
        'fem_hd_famhh', 'Female-headed family households with kids',
        'pai_hh',       'Households with public assistance income') %>% 
  kableExtra::kable() %>% 
  kableExtra::kable_minimal()
```


Two composite variables will be created from a subset of those listed above:  1) The count of residents living in the same home one year ago will be subtracted from the total population to create a `recent_movers` variable; and 2)
The multiple variables counting men between the ages of 18 and 29 will be summed to create a `young_males` variable.

For demonstration purposes, this script will use only two variables, the totals of residential population and households, in examples.


## API Call


Let's examine the structure of the API call.  It requires:

* The dataset year (we want `2010-2025`, downloaded one at a time)
* The survey name (`survey = 'acs5'`)
* A geographic level (`level = tract`)
* A named list of Census variables (where the names are the variable names we will use, and the values are the official names of the Census variables)
* The [FIPS codes](https://www.census.gov/programs-surveys/geography/guidance/geo-identifiers.html) for the state (`state = '17'`) and county (`county = '031'`)
* A Census API key, stored locally in a JSON file: `~/census_api_key.json`


```{r api_call_str}
'https://api.census.gov/data/{yr}/acs/{survey}?get=NAME,{str_c(l_vars, collapse = ",")}&for={level}:*&in=state:{state}%20county:{str_c(county, collapse = ",")}&key={jsonlite::read_json(PTH.CENSUS_API_KEY)$key}'
```


Once we paste the required elements into the string, we can then call the API. Here we test the call with `yr = 2010` and for the two variables specified above.

First we build the string, and then immediately call the API and store the raw result. Note that we will not examine the call string as it contains the API key, which should stay private.


```{r build_call}
#' Select sample year.
yr <- 2010

#' Abbreviated named list of Census variables (for examples).
l.vars <- list(pop.t = 'B01003_001E', # Total residential population
               hh.t  = 'B19057_001E') # Total households

#' Build API call; call the API; extract the content and store in a list.
l.content <- 
  glue('https://api.census.gov/data/{yr}/acs/{SURVEY}?get=NAME,',
       '{str_c(l.vars, collapse = ",")}&for={LEVEL}:*&in=state:{STATE}%20',
       'county:{str_c(COUNTY, collapse = ",")}&key={jsonlite::read_json(',
       'PTH.CENSUS_API_KEY)$key}') %>% 
  httr::GET() %>% 
  httr::content()
```


The API returns a list of lists, one for each tract in Cook County plus an additional row with variable headers. Here are the first two elements of the list, that is, the variable headers and the first tract observation:


```{r glance_content}
head(l.content, 2)
```


## Restructuring Content


We need to restructure the list into a dataframe.  We start by unlisting the top level only, meaning that every remaining element is a one-element list holding a value (either a column name or a value from any variable). We do this with `recursive = FALSE` to retain any `NULL` values, then convert those to `NA` and restructure the list into a vector, then into a matrix, and then into a tibble (assigning temporary numbered column names with `.name_repair = 'universal_quiet'`).  The column names are set to the values in the first row, and then that row is dropped.  The `GEOID` field is built from the `state`, `county`, and `tract` IDs; it and a differently cased `Name` field are both created to the right of the other variables. The original place ID and `NAME` variables are dropped.

At this point, the Census variables of interest will occupy the two leftmost columns, in the order in which they were requested by the API. We need to rename the Census variables to match the desired names specified by `l.vars` (specifically, by `names(l.vars)`). To do this, we concatenate the desired names with the existing variable names *minus* those with an underscore---these are the Census-assigned variable names that we want to replace. Now we convert the count variables to integers. If the dataset contained the requisite variables for `young_males` and `recent_movers`, we would create those with `f.census_composite_vars()`. Finally, the columns are reordered.


```{r api_content_to_df}
(df.census_cook.2010 <- 
  l.content %>% 
  unlist(recursive = FALSE) %>% 
  map_vec(~ ifelse(is.null(.), NA, .)) %>% 
  matrix(ncol  = length(l.vars) + 4,
         byrow = TRUE) %>%
  as_tibble(.name_repair = 'universal_quiet') %>%
  set_names(slice(., 1)) %>%
  slice(-1) %>% 
  mutate(GEOID = str_c(state, county, tract),
         Name  = NAME) %>%
  select(-c(NAME, state, county, tract)) %>% 
  set_names(c(names(l.vars),
              names(.) %>%
                keep(!str_detect(., '_')))) %>% 
  mutate(across(names(l.vars), as.integer)) %>% 
  # f.census_composite_vars() %>%
  select(GEOID, Name, everything())) %>% 
  head(3)
```


## Add Geometry


Next, we add Census-tract geometry for 2010.  (The `f.read_geo_tracts()` function is pretty simple, so I won't go through it line by line.)


```{r geo_cook_2010_glance_map}
#' Download tract geometry and merge into Census data. Preview, then map.
(geo.census_cook.2010 <- 
  f.read_geo_tracts(yr, STATE, COUNTY) %>% 
  select(-any_of(c('State', 'County', 'Name'))) %>% 
  full_join(df.census_cook.2010, by = 'GEOID')) %>% 
  print(n = 3) %>% 
  mapview(layer.name = 'Cook County - 2010')
```


## Download Full Dataset


We can download all fifteen years of Census datasets in one fell swoop with the `f.acs5_longitudinal()` function, saving it for future reference.


```{r download_full}
#' Download data from the Census API and save it.
tictoc::tic()
l.geo.census_cook <- f.acs5_longitudinal(V.YEARS_CENSUS, save = TRUE)
tictoc::toc()
```


In the next script, we'll limit the dataset to Chicago tracts and reshape the tracts so that the boundaries are consistent over time.