---
title: "04 - Download Crime Data"
output: html_document
---


This script downloads the necessary crime data from the [Chicago Data Portal](data.cityofchicago.org), then prepares the data for further exploration.

We will explore deprivation-of-property crimes taking place between 2010 and 2024. These are those with a primary type of burglary, deceptive practice (but only thefts by lessees or of mislaid property), motor vehicle theft, robbery, or theft (but non identity theft).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#' Load libraries, constants, and functions.
source('functions/f03c--data--crimes.R')

#' Download deprivation-of-property crimes, 2010-2024.
#' About 1-2 minutes.
tictoc::tic()
df.crime_raw <- f.crime_raw(download = TRUE, query = Q.CRIME)
tictoc::toc()

#' Chicago city boundaries.
geo.chi <-
  dir('data', 'geo--chicago-city', full.names = TRUE) %>%
  qs2::qs_read()
```


Having downloaded the dataset from the Chicago Data Portal's API, we'll prepare it for further analysis. First, we'll add part-of-date elements for year, quarter, month, and day-of-week.


```{r part_of_date}
#' Use the first ten observations for example code.
df.crime_raw %>% 
  head(10) %>% 
  mutate(year    = year(date),
         quarter = quarter(date),
         month   = month(date),
         weekday = wday(date),)
```


Now we want to do some minor recoding for the `description` variable. The goal is to group similar descriptions and use consistent terminology. Here are all seventy-two combinations of `type` and `desription` in the raw data:


```{r distinct_type_description}
df.crime_raw %>% 
  distinct(type, description = str_sub(description, 1, 50)) %>% 
  arrange(type, description) %>% 
  print(n = nrow(.))
```


After recoding (see the `f.recode_description()` function definition for details), there are now only twenty-five combinations:


```{r recode_description}
df.crime_raw %>% 
  distinct(type, description) %>% 
  f.recode_description() %>% 
  distinct(type, description) %>% 
  arrange(type, description) %>% 
  print(n = nrow(.))
```


We're also going to convert the data to spatial points so that we can map their locations (down to the block).  Let's apply the changes permanently.  Here's a map with a small sample of observations:


```{r geo_clean_crime}
#' Clean; convert to spatial points. Map a random sample of 500 observations.
(geo.crime <- 
  f.clean_crime(df.crime_raw) %>% 
  f.geo_crime()) %>% 
  sample_n(500) %>% 
  print(n = 3) %>% 
  mapview(cex        = 3,
          layer.name = 'Sample of Events',
          legend     = FALSE) +
  mapview(geo.chi,
          col.region = '#999',
          layer.name = 'Chicago City Boundaries',
          legend     = FALSE)
```


Finally, we'll save the dataset for use in future explorations.


```{r write_geo_crime}
#' Write to QS2 format with maximum compression.
f.write_qs2(geo.crime,
            'data',
            'geo--crime--dep-prop--2010-2024',
            n_threads = 10)
```